DSCI 445 - Algal Blooms

### 2020-12-10 paper.Rmd
Added in the where we got the data

With the increasing use of reservoirs for municipalities, the quality of the 
water is very important. Although the water gets treated before it gets to the 
tap, the quality of the water as it reaches the water treatment plant is very 
important regarding the cost to treat the water for municipal use, and algal 
blooms impact the quality of the water dramatically. Most municipal water is 
monitored, but not all of the data is publicly available and some isn’t 
monitored in situ - otherwise known as collected by hand. However, Landsat is a
series of satellites that image the Earth surface, in repeated intervals of 14 
days; that is to say Fort Collins gets re-imaged every 14 days. Landsat 5 dates
back to 1984, and is no longer taking images, while landsat 7 (1999) and landsat
8 (2013) still record images today. The landsat imagery allows for a larger 
sample of data to look at the trends of algal blooms in Colorado, in comparison 
to the lack of public in situ data available. Figure 1 shows reservoirs - only 
those greater than 0.01 km² -  that have both in situ and satellite imagery in 
blue, while the reservoirs in red indicate the reservoirs with just the landsat 
imagery available. It is easy to see the lack of reservoir data we have 
available to us, which is why utilizing machine learning to try and predict 
algae blooms from just remotely sensed imagery would be an incredibly useful
method of determining water quality in reservoirs.

@@

>>>>

With the increasing use of reservoirs for municipalities, the quality of the 
water is very important. Although the water gets treated before it gets to the 
tap, the quality of the water as it reaches the water treatment plant is very 
important regarding the cost to treat the water for municipal use, and algal 
blooms impact the quality of the water dramatically(Brookes). Most municipal water is 
monitored, but not all of the data is publicly available and some isn’t 
monitored in situ - otherwise known as collected by hand. However, Landsat is a
series of satellites that image the Earth surface, in repeated intervals of 14 
days; that is to say Fort Collins gets re-imaged every 14 days. Landsat 5 dates
back to 1984, and is no longer taking images, while landsat 7 (1999) and landsat
8 (2013) still record images today. The landsat imagery allows for a larger 
sample of data to look at the trends of algal blooms in Colorado, in comparison 
to the lack of public in situ data available. Figure 1 shows reservoirs - only 
those greater than 0.01 km² -  that have both in situ and satellite imagery in 
blue, while the reservoirs in red indicate the reservoirs with just the landsat 
imagery available. It is easy to see the lack of reservoir data we have 
available to us, which is why utilizing machine learning to try and predict 
algae blooms from just remotely sensed imagery would be an incredibly useful
method of determining water quality in reservoirs. It has been done previously,
but mainly utilizing only the blue and green wavelengths, despite there being more
potential options(Sachidananda).

@@

## Sources

Brezonik, Patrick, Kevin D. Menken, and Marvin Bauer. “Landsat-Based Remote
Sensing of Lake Water Quality Characteristics, Including Chlorophyll and Colored
Dissolved Organic Matter (CDOM).” Lake and Reservoir Management 21, no. 4 
(December 2005): 373–82. https://doi.org/10.1080/07438140509354442.

Brookes, J. D., and C. C. Carey. “Resilience to Blooms.” Science 334, no. 6052 
(October 7, 2011): 46–47. https://doi.org/10.1126/science.1207349.

Brookes, Justin D., Cayelan C. Carey, David P. Hamilton, Lionel Ho, Leon van der
Linden, Robert Renner, and Anna Rigosi. “Emerging Challenges for the Drinking 
Water Industry.” Environmental Science & Technology 48, no. 4 (February 18, 
2014): 2099–2101. https://doi.org/10.1021/es405606t.

Folke, Carl, Steve Carpenter, Brian Walker, Marten Scheffer, Thomas Elmqvist,
Lance Gunderson, and C.S. Holling. “Regime Shifts, Resilience, and Biodiversity 
in Ecosystem Management.” Annual Review of Ecology, Evolution, and Systematics 
35, no. 1 (December 15, 2004): 557–81. 
https://doi.org/10.1146/annurev.ecolsys.35.021103.105711.

Ho, Jeff C., Anna M. Michalak, and Nima Pahlevan. “Widespread Global Increase in
Intense Lake Phytoplankton Blooms since the 1980s.” Nature 574, no. 7780 
(October 2019): 667–70. https://doi.org/10.1038/s41586-019-1648-7.

Olmanson, Leif G, Marvin E Bauer, and Patrick L Brezonik. “Use of Landsat
Imagery to Develop a Water Quality Atlas of Minnesota’s 10,000 Lakes.” 
Conference Proceedings, 2002, 8.

Scheffer, Marten, Steve Carpenter, Jonathan A. Foley, Carl Folke, and Brian 
Walker. “Catastrophic Shifts in Ecosystems.” Nature 413, no. 6856 (October 2001)
591–96. https://doi.org/10.1038/35098000.

Torbick, Nathan, Sarah Hession, Stephen Hagen, Narumon Wiangwang, Brian Becker, 
and Jiaguo Qi. “Mapping Inland Lake Water Quality across the Lower Peninsula of
Michigan Using Landsat TM Imagery.” International Journal of Remote Sensing 34,
no. 21 (November 10, 2013): 7607–24. https://doi.org/10.1080/01431161.2013.822602.
